Modeling and analyzing the Corona-virus warning app with the Isabelle Infrastructure framework

### 2020-07-03 paper.tex
added a refinement section copied from arxive paper and needs to be  merged in. Also adapted the intro and background on Isabelle Infrastructures.

\newcommand{\symb}[1]{\makebox{\it #1}} 
\newcommand\ie{i.e.\!\,, }
\newcommand\rel{Re${\cal{L}}$}
\newcommand{\TODO}[1]{\textcolor{red}{\textbf{[TODO:#1]}}}
\newcommand\ttand{\mbox{{$\land$}}}
\newcommand\ttor{\mbox{{$\lor$}}}
\newcommand\ttcap{\mbox{{$\cap$}}}
\newcommand\ttcup{\mbox{{$\cup$}}}
\newcommand\ttfun{\mbox{{$\Rightarrow$}}}
\newcommand\ttmimp{\mbox{{$\Longrightarrow$}}}
\newcommand\ttimp{\mbox{{$\longrightarrow$}}}
\newcommand\ttequiv{\mbox{{$\equiv$}}}
\newcommand\ttexists{\mbox{{$\exists$}}}
\newcommand\ttforall{\mbox{{$\forall$}}}
\newcommand\ttneg{\mbox{{$\neg$}}}
\newcommand\ttneq{\mbox{{$\neq$}}}
\newcommand\ttin{\mbox{{$\in$}}}
\newcommand\ttnin{\mbox{{$\notin$}}}
\newcommand\ttImp{\mbox{{$\Longrightarrow$}}}
\newcommand\ttlam{\mbox{\( \lambda \)}}
\newcommand\tttimes{\mbox{\( \times \)}}
\newcommand\ttlbrack{\mbox{\(\llbracket\)}}
\newcommand\ttrbrack{\mbox{\( \rrbracket \)}}
\newcommand\noie{\textit{i.e.},\xspace}
\newcommand\eg{~\textit{e.g.},\xspace}
\newcommand\noeg{\textit{e.g.},\xspace}
\newcommand\ttatI{\mbox{\( @_G \)}}
\newcommand\ttto[1]{\mbox{{$\to^{#1}$}}}
\newcommand\ttleq{\mbox{{$\le$}}}
\newcommand\ttrelI{\mbox{{$\to_{i}$}}}
\newcommand\ttalpha{\mbox{{$\alpha$}}}
\newcommand\tttau{\mbox{{$\tau$}}}
\newcommand\ttsubseteq{\mbox{{$\subseteq$}}}
\newcommand\ttf{\mbox{{$f$}}}
\newcommand\ttvdash{\mbox{{$\vdash$}}}
\newcommand\ttref[1]{\mbox{\(\sqsubseteq_{#1}\)}}
\newcommand\ttNatt{\mbox{{$\mathcal{N}$}}}
\newcommand\ttattand[1]{\mbox{{$\oplus_{\wedge}^{#1}$}}}
\newcommand\ttattor[1]{\mbox{{$\oplus_{\vee}^{#1}$}}}
\newcommand\ttrelIstar{\mbox{{$\to_{i}^*$}}}

@@

In this paper, we first provide some background in Section \ref{sec:background}:
we give a detailed overview of the development history and security and privacy
relevant parts of the Corona-virus warning app (Section \ref{sec:history})
and some essential facts about the Isabelle Infrastructure framework
(Section \ref{sec:isainf}).
We then present our model (Section \ref{sec:model}) and analysis of privacy and
attacks (Section \ref{sec:ana}) before drawing some conclusions (Section \ref{sec:concl}).

@@

\subsection{Isabelle Infrastructure framework}
\label{sec:isainf}
\TODO{Adapt: this is copied from FMBC paper}
The Isabelle Infrastructure is built in the interactive generic theorem prover
Isabelle/HOL \cite{npw:02}. As a framework, it supports formalisation and proof of
systems with actors and policies. It originally emerged from verification of insider 
threat scenarios but it soon became clear that the theoretical concepts, like temporal
logic combined with Kripke structures and a generic notion of state transitions were
very suitable to be combined with attack trees into a formal security engineering process
\cite{suc:16} and framework \cite{kam:19a}.

Figure \ref{fig:theorystruc} gives an overview of the Isabelle Infrastructure 
framework with its layers of object-logics -- each level below embeds the one
above showing the novel contribution of this paper in blue on the top. 
\begin{figure}
\begin{center}
  \includegraphics[scale=.5]{theory_structure.pdf}
\end{center}
%\vspace{-.5cm}
\caption{Generic Isabelle Infrastructure framework applied to Corona-warning app.}
\label{fig:theorystruc}
\end{figure}
The formal model of the Corona-warning app uses the Isabelle Infrastructure framework instantiating it
by reusing its concept of {\it actors} for users and smartphones whereby locations correspond
to physical locations. The Ephermeral IDs, their sending and change is added to Infrastructures
by slightly adapting the basic state type of infrastructure graphs and accordingly the semantic rules
for the actions move, get, and put. The details of the newly adapted Infrastructure are
presented in Section \ref{sec:model}.
Technically, an Isabelle theory file \texttt{Infrastructure.thy} builds on top of the theories for Kripke 
structures and CTL (\texttt{MC.thy}), attack trees (\texttt{AT.thy}), and security refinement 
(\texttt{Refinement.thy}). Thus all these concepts can be used to specify the formal model
for IBC, express relevant and interesting properties and conduct interactive proofs (with the 
full support of the powerful and highly automated proof support of Isabelle).
The IBC theory itself is an adaptation of the Infrastructure theory of the 
Isabelle Infrastructure framework and reuses (or slightly adapts) existing concepts. 
In the remainder of this paper, we introduce the model that we conceived for IBC. 
All Isabelle sources are available online \cite{kam:20gitsc}.

@@

>>>>

\newcommand{\symb}[1]{\makebox{\it #1}} 
\newcommand\ie{i.e.\!\,, }
\newcommand\rel{Re${\cal{L}}$}
\newcommand{\TODO}[1]{\textcolor{red}{\textbf{[TODO:#1]}}}
\newcommand\ttand{\mbox{{$\land$}}}
\newcommand\ttor{\mbox{{$\lor$}}}
\newcommand\ttcap{\mbox{{$\cap$}}}
\newcommand\ttcup{\mbox{{$\cup$}}}
\newcommand\ttfun{\mbox{{$\Rightarrow$}}}
\newcommand\ttmimp{\mbox{{$\Longrightarrow$}}}
\newcommand\ttimp{\mbox{{$\longrightarrow$}}}
\newcommand\ttequiv{\mbox{{$\equiv$}}}
\newcommand\ttexists{\mbox{{$\exists$}}}
\newcommand\ttforall{\mbox{{$\forall$}}}
\newcommand\ttneg{\mbox{{$\neg$}}}
\newcommand\ttneq{\mbox{{$\neq$}}}
\newcommand\ttin{\mbox{{$\in$}}}
\newcommand\ttnin{\mbox{{$\notin$}}}
\newcommand\ttImp{\mbox{{$\Longrightarrow$}}}
\newcommand\ttlam{\mbox{\( \lambda \)}}
\newcommand\tttimes{\mbox{\( \times \)}}
\newcommand\ttlbrack{\mbox{\(\llbracket\)}}
\newcommand\ttrbrack{\mbox{\( \rrbracket \)}}
\newcommand\noie{\textit{i.e.},\xspace}
\newcommand\eg{~\textit{e.g.},\xspace}
\newcommand\noeg{\textit{e.g.},\xspace}
\newcommand\ttatI{\mbox{\( @_G \)}}
\newcommand\ttto[1]{\mbox{{$\to^{#1}$}}}
\newcommand\ttleq{\mbox{{$\le$}}}
\newcommand\ttrelI{\mbox{{$\to_{i}$}}}
\newcommand\ttrelIstar{\mbox{{$\to^*$}}}
\newcommand\ttrel[1]{\mbox{{$\to_{#1}$}}}
\newcommand\ttrelstar[1]{\mbox{{$\to_{#1}^*$}}}
\newcommand\ttalpha{\mbox{{$\alpha$}}}
\newcommand\ttsupseteq{\mbox{{$\supseteq$}}}
\newcommand\ttupdownarrow{\mbox{{$\Updownarrow$}}}
\newcommand\tttau{\mbox{{$\tau$}}}
\newcommand\ttsubseteq{\mbox{{$\subseteq$}}}
\newcommand\ttf{\mbox{{$f$}}}
\newcommand\ttvdash{\mbox{{$\vdash$}}}
\newcommand\ttref[1]{\mbox{\(\sqsubseteq_{#1}\)}}
\newcommand\ttNatt{\mbox{{$\mathcal{N}$}}}
\newcommand\ttattand[1]{\mbox{{$\oplus_{\wedge}^{#1}$}}}
\newcommand\ttattor[1]{\mbox{{$\oplus_{\vee}^{#1}$}}}
\newcommand\ttsigma{\mbox{{$\sigma$}}}
\newcommand\ttmref[1]{\mbox{{$\sqsubseteq_{#1}$}}}
\newcommand\ttmeref{\ttmref{\mathcal{E}}}
\newcommand\ttsn{\mbox{{\texttt{s}$_0$}}}
\newcommand\ttsigmap{\mbox{{$\sigma'$}}}
\newcommand\ttecal{\mbox{$\mathcal{E}$}}
\newcommand\ttimg{\mbox{$\triangleleft$}}

@@

The contributions of this paper are (a) formal re-engineering of the Corona-virus warning app
(b) providing an additional security and privacy analysis with interactice theorem proving
certification of a novel view on the system architecture including actors, locations, and policies,
(c) an formal definition of a security refinement process that allows to improve a system
based on attacks found by the attack tree analysis.

In this paper, we first provide some background in Section \ref{sec:background}:
we give a detailed overview of the development history and security and privacy
relevant parts of the Corona-virus warning app (Section \ref{sec:history})
and some essential facts about the Isabelle Infrastructure framework
(Section \ref{sec:isainf}).
We then present our model (Section \ref{sec:model}) and analysis of privacy and
attacks (Section \ref{sec:ana}). The found attck on the first abstract specification
motivates refinement. The formal definition of refinement for the Isabelle Infrastructure
framework is introduced and illustrated on the Corona-virus warning app (Section \ref{sec:ref})
before drawing some conclusions (Section \ref{sec:concl}).

@@

\subsection{Isabelle Infrastructure framework}
\label{sec:isainf}
%\TODO{Adapt: this is copied from FMBC paper}
The Isabelle Infrastructure is built in the interactive generic theorem prover
Isabelle/HOL \cite{npw:02}. As a framework, it supports formalisation and proof of
systems with actors and policies. It originally emerged from verification of insider 
threat scenarios but it soon became clear that the theoretical concepts, like temporal
logic combined with Kripke structures and a generic notion of state transitions were
very suitable to be combined with attack trees into a formal security engineering process
\cite{suc:16} and framework \cite{kam:19a}.

\subsubsection{Overview}
\label{sec:isainfover}
Figure \ref{fig:theorystruc} gives an overview of the Isabelle Infrastructure 
framework with its layers of object-logics -- each level below embeds the one
above showing the novel contribution of this paper in blue on the top. 
\begin{figure}
\begin{center}
  \includegraphics[scale=.5]{theory_structure.pdf}
\end{center}
%\vspace{-.5cm}
\caption{Generic Isabelle Infrastructure framework applied to Corona-warning app.}
\label{fig:theorystruc}
\end{figure}

\subsubsection{Instantiation of Framework}
The formal model of the Corona-warning app uses the Isabelle Infrastructure framework instantiating it
by reusing its concept of {\it actors} for users and smartphones whereby locations correspond
to physical locations. The Ephermeral IDs, their sending and change is added to Infrastructures
by slightly adapting the basic state type of infrastructure graphs and accordingly the semantic rules
for the actions move, get, and put. The details of the newly adapted Infrastructure are
presented in Section \ref{sec:model}.
Technically, an Isabelle theory file \texttt{Infrastructure.thy} builds on top of the theories for Kripke 
structures and CTL (\texttt{MC.thy}), attack trees (\texttt{AT.thy}), and security refinement 
(\texttt{Refinement.thy}). Thus all these concepts can be used to specify the formal model
for the Corona-virus warning app, express relevant and interesting properties, and conduct
interactive proofs (with the full support of the powerful and highly automated proof support
of Isabelle).
%The Corona-virus warning app theory itself is an adaptation of the Infrastructure theory of the 
%Isabelle Infrastructure framework and reuses (or slightly adapts) existing concepts. 
%In the remainder of this paper, we introduce the model that we conceived for . 
All Isabelle sources are available online \cite{kam:20gitsc}.

\subsubsection{Refinement}
An interesting feature that has been integrated into the Isabelle Infrastructure
framework motivated by security engineering formal specifications for IoT healthcare
system is  is an extension of the formal specifciation process introducing
refinement of Kripke structures \cite{kam:19a,kam:20a}. It refines a system model based on a 
formal definition of a combination of trace refinement and structural 
refinement (or datatype refinement). The definition allows to prove property preservation results 
crucial for an iterative development process.
The refinements of the system specification  can be interleaved with attack 
analysis while security properties can be proved in Isabelle. In each iteration
security qualities are accumulated while continuously attack trees scrutinize
the design.

@@

\section{Refinement}
\label{sec:ref}
\TODO{This section is just copied over from arxive paper - adapt}
Intuitively, a refinement changes some aspect of the type of
the state, for example, replaces a data type by a richer datatype or
restricts the behaviour of the actors. The former is expressed directly 
by a mapping of datatypes, the latter is incorporated into the state
transition relation of the Kripke structure that corresponds to the 
transformed model.
%This relationship between the Kripke structure can be better understood
%visually (see Figure \ref{fig:modtrans})
In other words, we can encode a refinement within our framework
as a relation on Kripke structures that is parametrized additionally by
a function that maps the refined type to the abstract type.
The direction ``from refined to abstract'' of this type mapping may seem 
curiously counter-intuitive. However, the actual refinement is given by the 
refinement that uses this function as an input. The refinement
then refines an abstract to a more concrete system specification. 
The additional layer for the refinement 
%(the red box in Figure \ref{fig:theorystruc})
can be formalised in Isabelle as a
%second\footnote{The first refinement 
%relation in this framework is on attack trees summarized in Section \ref{sec:at}.} 
refinement relation 
$\sqsubseteq_{\mathcal{E}}$. 
The relation \texttt{mod\_trans} is typed as a relation over triples --
a function from a threefold Cartesian product to \texttt{bool}, the 
type containing true and false only.  
The type variables $\sigma$ and $\sigma'$ input to the type constructor 
\texttt{Kripke} represent the abstract state type and the concrete state type. 
Consequently, the middle element of the triples selected by the relation 
\texttt{mod\_trans} is a function of type $\sigma' \Rightarrow \sigma$ 
mapping elements of the refined state to the abstract state.
The expression in quotation marks after the type is again the
infix syntax in Isabelle that allows the definition of mathematical notation
instead of writing \texttt{mod\_trans} in prefix manner. This nicer infix
syntax is already used in the actual definition.
Finally, the arrow \texttt{\ttImp} is the implication of Isabelle's  meta-logic
while $\ttimp$ is the one of the {\it object} logic HOL. 
They are logically equivalent but of different
types: within a HOL formula $P$, for example, as below $\forall x. P \ttimp Q$, only the implication $\ttimp$
can be used.
\begin{ttbox}
 mod_trans ::  (\ttsigma Kripke \tttimes (\ttsigma' \ttfun \ttsigma) \tttimes \ttsigma' Kripke) 
               \ttfun bool                  ("_ \ttmref{(\_)} _")
  K \ttmeref K' \ttequiv \ttforall s' \ttin states K'. \ttforall s \ttin init K'. 
             s \ttrelstar{\sigma'} s' \ttimp \ttecal(s) \ttin init K 
             \ttand \ttecal(s) \ttrelstar{\sigma} \ttecal(s')
\end{ttbox}
The definition of \texttt{K \ttmeref\, K'} states that for any state $s$ 
of the refined Kripke structure that can be reached by the state transition
in zero or more steps from an initial state $s_0$ of the refined Kripke 
structure, the mapping ${\mathcal E}$ from the refined to the abstract 
model's state must preserve this reachability, i.e., the image of
$s_0$ must also be an initial state and from there the image of $s$
under ${\mathcal E}$ must be reached with $0$ or $n$ steps.

\subsection{Property Preserving System Refinement}
A first direct consequence of this definition is the following lemma
where the operator \texttt{$\ttimg$} in \texttt{\ttecal\ttimg(init K')}
represents function image, that is the set, $\{\ttecal(x). x \in \texttt{init K'}\} $.
\begin{ttbox}
{\bf{lemma}} init_ref: K \ttmeref K' \ttImp \ttecal\ttimg(init K') \ttsubseteq init K
\end{ttbox}
A more prominent consequence of the definition of refinement 
is that of property preservation. Here, we show that refinement preserves the
CTL property of ${\sf EF} s$ which means that a reachability property true in the
refined  model \texttt{K'} %has also been 
is already true in the abstract model.
A state set $s'$ represents a property %since we use 
in the predicate transformer view of properties as sets of states. 
The additional condition on initial states ensures that we cannot ``forget'' them. 
%initial states.
%The theorem states that, if the 
%property \texttt{{\sf EF} s'} is true in the concrete model its  has been true 
%can be reached from the initial state in 
%\texttt{K'} then it is also possible  to reach the image of this property
% in the abstract Kripke structure \texttt{K}.
\begin{ttbox}
{\bf{theorem}} prop_pres: 
   K \ttmeref K'  \ttImp init K \ttsubseteq \ttecal\ttimg(init K') \ttImp
   \ttforall s' \ttin Pow(states K'). K' \ttvdash {\sf EF} s' 
              \ttimp K \ttvdash {\sf EF} (\ttecal\ttimg(s'))
\end{ttbox}
It is remarkable, that our definition of refinement by Kripke 
structure refinement entails property preservation and makes it possible 
to prove this as a theorem in Isabelle once for all, i.e., as a meta-theorem.
However, this is due to the fact that our generic definition of state transition
allows to explicitly formalise such sophisticated concepts like reachability.
For practical purposes, however, the proof obligation of showing that
a specific refinement is in fact a refinement is rather complex
justly because of the explicit use of the transitive closure of the state
transition relation.
In most cases, the refinement will be simpler. Therefore, we offer
additional help by the following theorem that uses a stronger characterisation
of Kripke structure refinement and shows that our refinement follows
from this.
\begin{ttbox}
{\bf{theorem}} strong_mt: 
\ttecal\ttimg(init K') \ttsubseteq init K \ttand s \ttrel{\sigma'} s' \ttimp \ttecal(s) \ttrel{\sigma} \ttecal(s') 
\ttImp K \ttmeref K'
\end{ttbox}
This simpler characterisation is in fact a stronger one: we could have $s \ttrel{\sigma'} s'$ 
in the refined Kripke structure \texttt{K'} and $\neg(\ttecal(s) \ttrel{\sigma} \ttecal(s'))$
but neither $s$ nor $s'$ are reachable from initial states in \texttt{K'}.
For cases, where we want to have the simpler one-step proviso but still need 
reachability we provide a slightly weaker version of \texttt{strong\_mt}.
\begin{ttbox}
{\bf{theorem}} strong_mt':  
\ttecal\ttimg(init K') \ttsubseteq init K \ttand (\ttexists s0 \ttin init K'. s0  \ttrelIstar s)
 \ttand s \ttrel{\sigma'} s' \ttimp \ttecal(s) \ttrel{\sigma} \ttecal(s') \ttImp K \ttmeref K'
\end{ttbox}

This idea of property preservation coincides with the classical idea of
trace refinement as it is given in process algebras like CSP. In this view,
the properties of a system are given by the set of its traces. Now, a refinement
of the system is given by another system that has a subset of the traces of the 
former one.
Although the principal idea is similar, we greatly extend it since our notion
additionally incorporates refinement. Since we include a state map 
\texttt{\ttsigma'\ttfun \ttsigma} in our refinement map, we additionally
allow structural refinement: the state map generalises the basic idea of
trace refinement by traces corresponding to each other but allows additionally
an exchange of data types. 
As we see in the application to the case study, the refinement steps may
sometimes just specialise the traces: in this case the state map 
\texttt{\ttsigma'\ttfun \ttsigma} is just identity.
